D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-16 cuttingroom.tex
Move additional material to cuttingroom.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

>>>>

\subsubsection{Executions Can Be Merged}
\label{sec:Executions Can Be Merged}

Consider this example, similar to that shown in
Section~\ref{sec:Semantic Dependencies Not Affected by if Statements},
which appears to be an \co{if} statement:

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}

Because the stores executed on each branch of that \co{if} statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should should err on the
side of caution by assuming there is no semantic
dependency~\cite{Howells2009membartxt},
the fact remains that reasonable C++ implementations might disagree as
to whether this code represents one execution or two.
It is the implementations's choice.

A similar caution applies to loads:

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 42)
    r2 = y;
else
    r2 = y;
z = r2;
\end{verbatim}
\end{quote}

The two loads from \co{y} are fungible, and so a C++ implementation
could behave as if this \co{if} statement had instead been a single load
from \co{y}.

These examples demonstrate a key point: 
\emph{In general, executions do not necessarily map one-to-one onto
paths through the source code}.% \footnote{
% 	Thanks to Alan Stern for the examples in this section and
% 	in the preceding one.}

\subsubsection{Executions Depend on Context}
\label{sec:Executions Depend on Context}

Consider the following example derived from that in
Section~\ref{sec:Semantic Dependencies and Source Code}:

\begin{quote}
\begin{verbatim}
r1 = x;
// r1 = 0;
z = y * r1;
// z = y / r1;
\end{verbatim}
\end{quote}

As written, there is a semantic dependency from \co{x} and \co{y}
to \co{z}.
But if the second line is uncommented, there is no semantic
dependency.
If the second line is left commented, but the last line is uncommented,
then the compiler is free to backwards-propagate the potential
divide-by-zero undefined behavior to precede the first assignment
to \co{z}.
This could prevent both assignments to \co{z} from executing, for example,
due to a divide-by-zero exception.

More subtle changes in context can also have an effect.
For example:

\begin{quote}
\begin{verbatim}
r1 = x;
z = y * r1;
z = 42;
\end{verbatim}
\end{quote}

Given that there is no ordering, a C++ implementation could choose to
act as if the first store to \co{z} was omitted, even if that store was
a relaxed atomic store.
But if \co{z} was marked volatile, both stores would need to be executed
in strict accordance with the rules of the abstract machine.

\emph{Any analysis of a given execution either must be fully informed
of the enclosing context, or must make conservative assumptions about
that context.}

\subsubsection{Executions Can Be Nondeterministic}
\label{sec:Executions Can Be Nondeterministic}

Consider this example, similar to that in
Section~\ref{sec:Abstract Executions},
with \co{i} initially zero:

\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

y = foo(++i, ++i);
x = y * z;
\end{verbatim}
\end{quote}

Because early C~implementers could not come to agreement, standard
does not specify the order of evaluation of function arguments, so the
resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
($\frac{2}{1}$).
In the former case, there is no semantic dependency from \co{z} to
\co{x}, but in the latter case there is.
Thus, semantic dependencies are not just a function of a particular
execution through the source code, but also of arbitrary choices made
by the C++ implementation.\footnote{
	Again, thanks to Peter Sewell for pointing out this possibility.}

Note well that C++ implementations can and do act as if portions of
the source code was duplicated in order to apply specialization
optimizations, in which case any non-determinism might also be
a function of a given execution.

Many projects' coding guidelines prohibit side effects in expressions,
their goal being to obtain portable code whose behavior does not depend
on arbitrary choices on the part of the implementation.
However, the above code nevertheless conforms to the standard.
Note that side effects in expressions lacking sequence points is
undefined behavior, however, there are proposals to define it in the
C++~\cite{GabrielDosReis2016P0145r3} and
C~\cite{AlexCeleste2023N3203}
languages, allowing implementations to nevertheless reorder
evaluation when permitted by the as-if rule.

Even though the standard is non-deterministic, \emph{compiler-based C++
implementions make their non-deterministic choices at compile time, so
that the object code is deterministic, at least on a per-execution basis.}

\subsubsection{Implementers and Users Influence the Definition of Execution}
\label{sec:Implementers and Users Influence the Definition of Execution}

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
Users  and implementers are of course wise to assume that their opinions
might be overridden by those of the standard and, for the users,
especially the implementation.
However, consider the following code, similar to that shown in
Section~\ref{sec:Abstract Executions}:

\begin{quote}
\begin{verbatim}
void foo(int x, char c)
{
    return x * (c >= 0);
}
\end{verbatim}
\end{quote}

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above has a semantic dependency in executions running
on systems choosing signed \co{char}, but not those running on systems
choosing unsigned \co{char}.
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

Except that GCC provides the \co{-funsigned-char} command-line
argument that causes this implementation to treat variables of
type \co{char} as unsigned, regardless of the architects' wishes.
Given this command-line argument, there is always a semantic dependency
in the execution running from the \co{foo()} function's argument to its
return value.
So score one for at least some implementations' users.

Please see Appendix~\ref{app:User Influence Over Language Semantics}
for more examples of user control over semantics, including semantic
dependencies.

\emph{A definition of ``execution'' drawn strictly from the standard will
be at best an approximation to a definition that is useful in practice.}

\subsubsection{So What Exactly is an Execution, Anyway?}
\label{sec:So What Exactly is an Execution, Anyway?}

Given all of these complications in C++ executions, and given that
semantic dependencies must be defined in terms of executions, it makes
sense to lower our sights from full generality.
We therefore restrict ourselves to a specific context:
traditional compilers.
It will then turn out to be necessary to take the notion of semantic
dependence, and thus OOTA, to be compiler-relative rather than absolute.

We will show that programs built using loose C++ compilers that treat all
atomic operations as volatile can never exhibit OOTA, and the same is true
for compilers obeying the weaker constraint that atomic accesses may not
invented, provided the compiler does not perform multi-thread analysis.

% @@@ LTO?  Does not appear to be relevant given new material.

These restrictions permit us to bring to bear the real-world
constraints introduced by the next section.

### 2024-02-16 ootavacuous.tex
Move additional material to cuttingroom.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\section{To Be Moved Or Removed...}

@@@ Review

\subsubsection{Executions Can Be Merged}
\label{sec:Executions Can Be Merged}

Consider this example, similar to that shown in
Section~\ref{sec:Semantic Dependencies Not Affected by if Statements},
which appears to be an \co{if} statement:

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}

Because the stores executed on each branch of that \co{if} statement store
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
(for all intents and purposes) a single store.
Although software projects relying on ordering from load-to-store control
dependencies and using a variety of compilers should should err on the
side of caution by assuming there is no semantic
dependency~\cite{Howells2009membartxt},
the fact remains that reasonable C++ implementations might disagree as
to whether this code represents one execution or two.
It is the implementations's choice.

A similar caution applies to loads:

\begin{quote}
\begin{verbatim}
r1 = x;
if (r1 > 42)
    r2 = y;
else
    r2 = y;
z = r2;
\end{verbatim}
\end{quote}

The two loads from \co{y} are fungible, and so a C++ implementation
could behave as if this \co{if} statement had instead been a single load
from \co{y}.

These examples demonstrate a key point: 
\emph{In general, executions do not necessarily map one-to-one onto
paths through the source code}.% \footnote{
% 	Thanks to Alan Stern for the examples in this section and
% 	in the preceding one.}

\subsubsection{Executions Depend on Context}
\label{sec:Executions Depend on Context}

Consider the following example derived from that in
Section~\ref{sec:Semantic Dependencies and Source Code}:

\begin{quote}
\begin{verbatim}
r1 = x;
// r1 = 0;
z = y * r1;
// z = y / r1;
\end{verbatim}
\end{quote}

As written, there is a semantic dependency from \co{x} and \co{y}
to \co{z}.
But if the second line is uncommented, there is no semantic
dependency.
If the second line is left commented, but the last line is uncommented,
then the compiler is free to backwards-propagate the potential
divide-by-zero undefined behavior to precede the first assignment
to \co{z}.
This could prevent both assignments to \co{z} from executing, for example,
due to a divide-by-zero exception.

More subtle changes in context can also have an effect.
For example:

\begin{quote}
\begin{verbatim}
r1 = x;
z = y * r1;
z = 42;
\end{verbatim}
\end{quote}

Given that there is no ordering, a C++ implementation could choose to
act as if the first store to \co{z} was omitted, even if that store was
a relaxed atomic store.
But if \co{z} was marked volatile, both stores would need to be executed
in strict accordance with the rules of the abstract machine.

\emph{Any analysis of a given execution either must be fully informed
of the enclosing context, or must make conservative assumptions about
that context.}

\subsubsection{Executions Can Be Nondeterministic}
\label{sec:Executions Can Be Nondeterministic}

Consider this example, similar to that in
Section~\ref{sec:Abstract Executions},
with \co{i} initially zero:

\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

y = foo(++i, ++i);
x = y * z;
\end{verbatim}
\end{quote}

Because early C~implementers could not come to agreement, standard
does not specify the order of evaluation of function arguments, so the
resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
($\frac{2}{1}$).
In the former case, there is no semantic dependency from \co{z} to
\co{x}, but in the latter case there is.
Thus, semantic dependencies are not just a function of a particular
execution through the source code, but also of arbitrary choices made
by the C++ implementation.\footnote{
	Again, thanks to Peter Sewell for pointing out this possibility.}

Note well that C++ implementations can and do act as if portions of
the source code was duplicated in order to apply specialization
optimizations, in which case any non-determinism might also be
a function of a given execution.

Many projects' coding guidelines prohibit side effects in expressions,
their goal being to obtain portable code whose behavior does not depend
on arbitrary choices on the part of the implementation.
However, the above code nevertheless conforms to the standard.
Note that side effects in expressions lacking sequence points is
undefined behavior, however, there are proposals to define it in the
C++~\cite{GabrielDosReis2016P0145r3} and
C~\cite{AlexCeleste2023N3203}
languages, allowing implementations to nevertheless reorder
evaluation when permitted by the as-if rule.

Even though the standard is non-deterministic, \emph{compiler-based C++
implementions make their non-deterministic choices at compile time, so
that the object code is deterministic, at least on a per-execution basis.}

\subsubsection{Implementers and Users Influence the Definition of Execution}
\label{sec:Implementers and Users Influence the Definition of Execution}

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
Users  and implementers are of course wise to assume that their opinions
might be overridden by those of the standard and, for the users,
especially the implementation.
However, consider the following code, similar to that shown in
Section~\ref{sec:Abstract Executions}:

\begin{quote}
\begin{verbatim}
void foo(int x, char c)
{
    return x * (c >= 0);
}
\end{verbatim}
\end{quote}

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above has a semantic dependency in executions running
on systems choosing signed \co{char}, but not those running on systems
choosing unsigned \co{char}.
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

Except that GCC provides the \co{-funsigned-char} command-line
argument that causes this implementation to treat variables of
type \co{char} as unsigned, regardless of the architects' wishes.
Given this command-line argument, there is always a semantic dependency
in the execution running from the \co{foo()} function's argument to its
return value.
So score one for at least some implementations' users.

Please see Appendix~\ref{app:User Influence Over Language Semantics}
for more examples of user control over semantics, including semantic
dependencies.

\emph{A definition of ``execution'' drawn strictly from the standard will
be at best an approximation to a definition that is useful in practice.}

\subsubsection{So What Exactly is an Execution, Anyway?}
\label{sec:So What Exactly is an Execution, Anyway?}

Given all of these complications in C++ executions, and given that
semantic dependencies must be defined in terms of executions, it makes
sense to lower our sights from full generality.
We therefore restrict ourselves to a specific context:
traditional compilers.
It will then turn out to be necessary to take the notion of semantic
dependence, and thus OOTA, to be compiler-relative rather than absolute.

We will show that programs built using loose C++ compilers that treat all
atomic operations as volatile can never exhibit OOTA, and the same is true
for compilers obeying the weaker constraint that atomic accesses may not
invented, provided the compiler does not perform multi-thread analysis.

% @@@ LTO?  Does not appear to be relevant given new material.

These restrictions permit us to bring to bear the real-world
constraints introduced by the next section.

>>>>

Despite these outstanding issues, our definition of semantic dependency
allows us to bring to bear the real-world constraints outlined in the
next section.
